Pythran: Enabling Static Optimization of Scientific Python Programs

### 2014-06-27 doc/papers/iop2014/paper.tex
Update article

    Pythran is an open source static compiler that turns modules written
    in a subset of Python language into native ones. Assuming that scientific
    modules do not rely much on the dynamic features of the language, it trades
    them for powerful, eventually inter procedural, optimizations.
    These optimizations include detection of pure functions, temporary
    allocation removal,
    constant folding, Numpy \texttt{ufunc} fusion and parallelization, explicit
    thread-level parallelism through OpenMP annotations, false variable
    polymorphism pruning, and automatic vector instruction generation such as
    AVX or SSE.

@@

    % FIXME  (explain that annotations can be added to the code?)
    Unlike the Cython approach , Pythran input code remains compatible with the
    Python interpreter. Output code is generally as efficient as the annotated
    Cython equivalent, if not more, but without the backward compatibility
    loss.

@@

The Python language is growing in popularity as a language for scientific
computing, thanks to % FIXME (sounds strange for a publication)
 its concise syntax, a high level standard library and
several scientific packages.

However, the overhead of running a scientific application written in Python
compared to the same algorithm written in a statically compiled language such
as C is high. % FIXME cut sentence It is partly due to the interpration cost,
but also to Python's dynamic binding. Additionally, the Python compiler does
not perform any optimization on the byte code, while scientific applications
are first-class candidates for many of them.

Since scientific applications are often told to spend 90\% of their time in
10\% of the code, it is natural to focus on those computation-intensive pieces 
of code. The goal of an optimizing compiler % FIXME (of what) 
may not be to optimize the full Python application, but rather a subset of the application.

@@

Except for PyPy, none of these compilers apply any of the static optimization
techniques that have been known and successfully applied 
% FIXME And Numba? It generates LLVM, which uses static optimization techniques
for decades to statically compiled language such as C or C++.  Translators to
statically compiled languages do take advantage of them indirectly, but the
quality of the generated code may prevent advanced optimizations. For instance
GCC hardly vectorizes the output of Cython, while the array semantic is
avaialable at the Python level. Taking into account the specificities of the
Python language can unlock many new transformations.  For instance, PyPy
automatically converts the \texttt{range} builtin into \texttt{xrange} by using
a dedicated structure called \texttt{range-list}.

@@

Finally, Pythran computes the set of functions and types used for the processed
module, and generates the minimal set of headers that need to be included to
provide the definitions needed to compile the generated C++ code. This step is
critical to prevent typical slowdown during C++ compilation.

The second step only consists in the instantiation of the top-level functions of
the module, using user-provided signatures. Template instantiation then triggers
the different correctly typed instantiations for all functions called in the
module. Note that the user only needs to provide the type of the functions
exported outside the module. The possible types of all internal functions are
then inferred from the call sites. Note that this complex step leverage the 
C++11 compiler capabilities: Pythran generates only one line using user-provided
signature to trigger the template instanciation by the C++ compiler.

The last step involves a template library, called \texttt{pythonic} that 
contains a polymorphic implementation of many functions of the Python standard
library in the form of C++ template functions. Several optimizations, most
notably using expression template, are delegated to this library. Pythran relies
on the C++11~\cite{isocxx11} language and leverage recent features such as
move semantics, type inference through \texttt{decltype(\dots)} and variadic
templates.
As a consequence it requires a C++11 compatible compiler for the native code
generation. Boost.Python~\cite{boostpython2007} is used for the automatic
Python-to-C++ boilerplate. Generated C++ code is compatible starting with g++ 
4.8.2 and clang++ 3.3.

It is important to note that all Pythran analyses and optimizations are
type-agnostic, \ie{they do not assume any type for the variables manipulated
by the program}. The generated C++ template is still type agnostic, the specialization occurs only by generating the template instanciation.
In other words, the Pythran compiler analyzes polymorphic functions and turns
Python modules into C++ meta-programs.

@@

A rather high-level analysis is the \texttt{PureFunctions} analysis, that computes the
set of functions declared in the module that are pure, \ie{whose return value
only depends from the value of their arguments and do not change any global state}.
This analysis depends on two
other analyses, namely \texttt{GlobalEffects} that computes for each function whether
this function modifies the global state (including I/O, random generators, etc.)
and \texttt{ArgumentEffects} that computes for each argument of each function 
whether or not this argument may be updated in the function body. These three
analyses work inter-procedurally, as illustrated in the following example:

@@

Next section performs an in-depth comparison of Pythran with others Python
optimizers: Cython and PyPy.

@@

All benchmarks presented in this section are run on an hyper-threaded quad-core
i7, % FIXME sandy-brige, nehalem, what?
using examples shipped along with
Pythran sources, available at \url{https://github.com/serge-sans-paille/pythran} in
the \texttt{pythran/test/cases} directory. The Pythran version used is the
\texttt{HEAD} of the \texttt{iop2014} branch, PyPy 2.0.2 compiled with the
\texttt{-jit} flag, CPython 2.7.3 and Cython 0.20. All timings are made using
the \texttt{timeit} module, taking the best of all runs. All C++ codes are
compiled with g++ 4.8.2, using the tool default compiler option, generally
\texttt{-O3} plus a few optimizing flags depending on the target. To make a
fair comparison, no parallelism annotations are added to any code.

@@

The ``rosen der'' kernel computes a complex numpy expression that accesses
data through a one dimensional slice. It is interesting to evaluate how much of
the expression can be compiled to regular, efficient loop while avoiding
temporaries.  Table~\ref{tbl:rosen} illustrate the benchmark results for
CPython, Cython, PyPy and Pythran. Cython slightly outperforms Pythran on that
test case, but Cython code had to be manually transformed into an explicit
loop. Moreover, activating automatic parallelization for Pythran yields an
additional $\times4$ speedup without a change to the input, while Cython needs
manual code annotation to achieve a similar result. It seems that PyPy support
for numpy expression is still note mature for numpy expression, but there is a
strong on-going effort of the community to fix this.

@@

The ``growcut'' application is an image processing code where all array indexing
are explicit. In that sense, this benchmark is complementary of the previous
one, as it evaluates how well explicit array access are compiled.
Table~\ref{tbl:growcut} summarizes the results. All compiler yield impressive
speedups on that benchmark, but this is is merely due to the fact that
accessing individual elements is extremely slow in numpy, and the direct C
translation of the Python code is very close to what a C developer would write.
Pythran appears to outperforms Cython once all access are explicit, most
certainly due to cleaner code generation.

@@

@@

Although Pythran focuses on a subset of Python and its standard library, many
optimizations opportunities are still available. Using a Domain Specific
Language (DSL) approach, one could use rewriting rules to optimize several
Python idioms. For instance, \texttt{len(set(x))} could lead to an optimized
\texttt{count\_uniq} that would iterate only once on the input sequence.

There is naturally more work to be done at the Numpy level, for instance to
support more functions from the original module. The extraction of Numpy
expressions from \texttt{for loops} is also a natural optimization candidate, which
shares similarities with code refactoring.

@@

To conclude, limiting Python to a statically typed subset does not hinders the
expressiveness when it comes to scientific or mathematic computations, but makes
it possible to use a wide variety of classical optimizations to help Python
match the performances of statically compiled language. Moreover, one can use
high level information to generate efficient code that would be difficult to
write for the average programmer.

>>>>

    Pythran is an open source static compiler that turns modules written in a
    subset of Python language into native ones. Assuming that scientific
    modules do not rely much on the dynamic features of the language, it trades
    them for powerful, possibly inter procedural, optimizations.  These
    optimizations include detection of pure functions, temporary allocation
    removal, constant folding, Numpy \texttt{ufunc} fusion and parallelization,
    explicit thread-level parallelism through OpenMP annotations, false
    variable polymorphism pruning, and automatic vector instruction generation
    such as AVX or SSE.

@@

    % FIXME  (explain that annotations can be added to the code?)
    Unlike the Cython approach, Pythran input code remains compatible with the
    Python interpreter. Output code is generally as efficient as the annotated
    Cython equivalent, if not more, but without the backward compatibility
    loss.

@@

The Python language is growing in popularity as a language for scientific
computing, thanks to % FIXME (sounds strange for a publication) its concise
syntax, a high level standard library and several scientific packages. The
combination of a relatively efficient array package --\texttt{numpy}--, a
library for scientific computing --\texttt{scipy}--, and enhanced interactive
console --\texttt{ipython}-- and a comprehensive 2D plotting package
--\texttt{matplotlib}-- makes it a relevant candidate for many daily scientific
tasks.

However, the overhead of running a scientific application written in Python
compared to the same algorithm written in a statically compiled language such
as C is high. It is partly due to the interpretation cost, but also to Python's
dynamic binding. Additionally, the Python compiler does not perform any
optimization on the byte code, while scientific applications are first-class
candidates for many of them. The ratio between an efficient C application and
its Python counterpart can range from a factor of two, when the Python code
only calls C routine, to a factor of one hundred or more when the Python code
makes explicit array accesses.

Given the increasing amount of data that scientists may need to process for
their experiments, even two can be a limiting factor for the adoption of Python
for computation intensive experiments. But since scientific applications are
often told to spend 90\% of their time in 10\% of the code, one can focus on
those computation-intensive pieces of code. The goal of an optimizing compiler
may not be to optimize the full Python application, but rather a subset of the
application.

@@

Except for PyPy, none of these compilers apply any of the static optimization
techniques that have been known and successfully applied 
% FIXME And Numba? It generates LLVM, which uses static optimization techniques
for decades to statically compiled language such as C or C++.  Translators to
statically compiled languages do take advantage of them indirectly, but the
quality of the generated code may prevent advanced optimizations. For instance
GCC hardly vectorizes the output of Cython, while the array semantic is
available at the Python level. Taking into account the specificities of the
Python language can unlock many new transformations.  For instance, PyPy
automatically converts the \texttt{range} builtin into \texttt{xrange} by using
a dedicated structure called \texttt{range-list}.

@@

Finally, Pythran computes the set of functions and types used for the processed
module, and generates the minimal set of headers that provide the definitions
needed to compile the generated C++ code. This step is critical to prevent
typical slowdown during C++ compilation.

The second step only consists in the instantiation of the top-level functions
of the module, using user-provided signatures. Template instantiation then
triggers the different correctly typed instantiations for all functions called
in the module. Note that the user only needs to provide the type of the
functions exported outside the module. The possible types of all internal
functions are then inferred from the call sites. This complex step leverage the
C++11 compiler capabilities: Pythran generates only one line using
user-provided signature to trigger the template instanciation by the C++
compiler.

The last step involves a template library, called \texttt{pythonic} that 
contains a polymorphic implementation of many functions of the Python standard
library in the form of C++ template functions. Several optimizations, most
notably using expression templates, are delegated to this library. Pythran relies
on the C++11~\cite{isocxx11} language and leverages recent features such as
move semantics, type inference through \texttt{decltype(\dots)} and variadic
templates.
As a consequence it requires a C++11 compatible compiler for the native code
generation. Boost.Python~\cite{boostpython2007} is used for the automatic
Python-to-C++ boilerplate. Generated C++ code is compatible starting with g++ 
4.8.2 and clang++ 3.3.

It is important to note that all Pythran analyses and optimizations are
type-agnostic, \ie{they do not assume any type for the variables manipulated
by the program}. The specialization of the generated code occurs only by generating the template instanciation.
In other words, the Pythran compiler analyzes polymorphic functions and turns
Python modules into C++ meta-programs.

@@

A rather high-level analysis is the \texttt{PureFunctions} analysis, that computes the
set of functions declared in the module that are pure, \ie{whose return value
only depends from the value of their arguments and do not change any global state}.
This analysis depends on two
other analyses, namely \texttt{GlobalEffects} that computes for each function whether
it modifies the global state (including I/O, random generators, etc.)
and \texttt{ArgumentEffects} that computes for each argument of each function 
whether or not this argument may be updated in the function body. These three
analyses work inter-procedurally, as illustrated in the following example:

@@

Next section performs an in-depth comparison of Pythran with three other Python
optimizers: CPython, Cython and PyPy.

@@

All benchmarks presented in this section are run on an hyper-threaded quad-core
i7, % FIXME sandy-brige, nehalem, what?
using examples shipped along with
Pythran sources, available at \url{https://github.com/serge-sans-paille/pythran} in
the \texttt{pythran/test/cases} directory. The Pythran version used is the
\texttt{HEAD} of the \texttt{iop2014} branch. The other tools are: PyPy 2.0.2 compiled with the
\texttt{-jit} flag, CPython 2.7.3 and Cython 0.20. All timings are made using
the \texttt{timeit} module, taking the best of all runs. All C++ codes are
compiled with g++ 4.8.2, using the tool default compiler option, generally
\texttt{-O3} plus a few optimizing flags depending on the target. To make a
fair comparison, no parallelism annotations are added to any code.

@@

The ``rosen der'' kernel computes a complex numpy expression that accesses
data through a one dimensional slice. It is interesting to evaluate how much of
the expression can be compiled to regular, efficient loop while avoiding
temporaries.  Table~\ref{tbl:rosen} illustrate the benchmark results for
CPython, Cython, PyPy and Pythran. Cython slightly outperforms Pythran on that
test case, but Cython code had to be manually transformed into an explicit
loop. Moreover, activating automatic parallelization for Pythran yields an
additional $\times4$ speedup without a change to the input, while Cython needs
manual code annotation to achieve a similar result. It seems that PyPy support
for numpy expression is still not mature for numpy expression, but there is a
strong on-going effort of the community to fix this.

@@

The ``growcut'' application is an image processing code where all array indexing
is explicit. In that sense, this benchmark is complementary to the previous
one, as it evaluates how well explicit array accesses are compiled.
Table~\ref{tbl:growcut} summarizes the results. All compilers yield impressive
speedups on that benchmark, but this is is merely due to the fact that
accessing individual elements is extremely slow in numpy, and the direct C
translation of the Python code is very close to what a C developer would write.
Pythran appears to outperforms Cython once all accesses are explicit, most
certainly due to cleaner code generation.

@@

\section{Integration into Existing Applications}

The \texttt{distutils} is the standard way to package applications. In addition
to its command line interface, Pythran provides a \texttt{distutils} extension
to integrate Pythran extension into existing applications. Its use is similar
to the \texttt{Extension} class:

\begin{lstlisting}
from distutils.core import setup
from pythran.dist import PythranExtension

module1 = PythranExtension('demo', sources = ['a.py'])

setup(name = 'demo',
      version = '1.0',
      description = 'This is a demo package',
      ext_modules = [module1])
\end{lstlisting}

As Pythran input language is a subset of Python, it is still possible to write
the \texttt{setup.py} so that the Python code is used as a regular module if
the Pythran import fails.

Pythran is supported on Linux and MacOSX systems.  It is available in source
code form on the git repository
\url{https://github.com/serge-sans-paille/pythran}, and on the cheese-shop
\url{https://pypi.python.org/pypi/pythran}.

@@

Although Pythran focuses on a subset of Python and its standard library, many
optimization opportunities are still available. Using a Domain Specific
Language (DSL) approach, one could use rewriting rules to optimize several
Python idioms. For instance, \texttt{len(set(x))} could lead to an optimized
\texttt{count\_uniq} that would iterate only once on the input sequence.

There is naturally more work to be done at the Numpy level, for instance to
support more functions from the original module. The extraction of Numpy
expressions from \texttt{for} loops is also a natural optimization candidate, which
shares similarities with code refactoring.

@@

To conclude, limiting Python to a statically typed subset does not hinders the
expressiveness when it comes to scientific or mathematic computations, but
makes it possible to use a wide variety of classical optimizations to help
Python match the performances of statically compiled language. Moreover, one
can use high level information to generate efficient code that would be
difficult to write for the average programmer. In the end, the scientific
community can benefit from a high-level language to quickly write their
experiments, and still run them at decent speed.
